Assessing spatial correlations in pulsar timing arrays with the noise-marginalized optimal statistic

### 2018-03-19 manuscript/manuscript.tex
updated description of simulated data

We used these three techniques to compute the optimal statistic for simulated data sets. 
We simulated PTA data sets with comprised of 18 MSPs 
with the same sky locations, observation times, and noise properties 
as the MSPs in the NANOGrav 9-year data set 
with at least 3 years of observation time \citep{abb+15} 
(see Table~\ref{tab:sim} for more details). 
To these data sets we injected an isotropic stochastic GWB 
\sv{insert strength of GWB... are we going to have one value or more than one?}.
\begin{table}[htb]
	\setlength{\tabcolsep}{5pt}
	\caption{Pulsar parameters used in simulated PTAs}
	\begin{center}
	\begin{tabular}{ccccc}
		\hline\hline
    		Pulsar	& $T_\mathrm{obs}$ (yrs) & $\sigma_w$ ($\mu\mathrm{s}$) & $\log_{10}A_\mathrm{red}$ & $\gamma_\mathrm{red}$ \\
		\hline
		J0030+0451 & & & \\
		J0613$-$0200 & 16.054 & 1.58 & -13.90 & 3.18 \\
		J1012+5307 & 16.831 & 1.47 & -13.09 & 1.65 \\
		J1024$-$0719 & & & \\
		J1455$-$3330 & & & \\
		J1600$-$3053 & & & \\
		J1614$-$2230 & & & \\
		J1640+2224 & 16.735 & 1.99 & -13.24 & 0.03 \\
		J1713+0747 & 17.657 & 0.26 & -14.90 & 4.85 \\
		J1741+1351 & & & \\
		J1744$-$1134 & 17.250 & 0.65 & -13.60 & 2.00 \\
		B1855+09 & & & \\
		J1853+1303 & & & \\
		J1909$-$3744 & 9.379 & 0.12 & -13.99 & 2.06 \\
		J1918$-$0642 & & & \\
		J2010$-$1323 & & & \\
		J2145$-$0750 & 17.161 & 1.19 & -13.87 & 4.02 \\
		J2317+1439 & & & \\
    		\hline\hline
	\end{tabular}
	\end{center}
	\label{tab:sim}
\end{table}

@@

In Fig.~\ref{fig:os_datasetstats} we compare the fixed-noise values of $\hat{A}^2$ 
to the mean $\hat{A}^2$ for 425 different realizations of the stochastic background. 
The fixed-noise analysis tends to underestimate the strength of the stochastic background, 
whereas the noise-marginalized analysis is better able to recover the injected value. 
The mean value of the optimal statistic from the fixed-noise analysis, 
averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
while the mean value from the noise-marginalized analysis is $\hat{A}^2 = 2.5 \times 10^{-29}$. 
The noise-marginalized analysis also gives a higher average mean SNR of 1.7, 
compared to an average SNR of 1.0 using the fixed-noise analysis.
%\begin{figure}[ht]
%	\includegraphics[width=0.9\columnwidth]{plots/os_datasetstats_5e-15.pdf}
%	\caption{Optimal statistic and SNR for 425 simulated datasets 
%			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
%			We compare the values found using a fixed-noise analysis (blue) to the 
%			mean optimal statistic and mean SNR found by marginalizing over 
%			the individual pulsars' red noise parameters. The dashed vertical line 
%			indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
%			The value of $\hat{A}^2$ from the fixed-noise analysis, 
%			averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
%			while the noise-marginalized analysis gives a mean value of $\hat{A}^2 = 2.5 \times 10^{-29}$. 
%			Furthermore, the fixed-noise analysis gives an average SNR of 1.0, while the 
%			noise-marginalized analysis gives an average mean SNR of 1.7.}
%	\label{fig:os_datasetstats}
%\end{figure}

>>>>

We used these three techniques to compute the optimal statistic 
for simulated ``NANOGrav-like'' data sets. 
These consisted of 18 MSPs with observation times up to 11 years 
and sky positions matching the 18 longest-observed pulsars in the 
NANOGrav 11-year data set \citep{abb+17}.
with the observation times increased so that the longest-observed pulsars 
had TOAs spanning over 11 years. 
We include white noise for all pulsars, plus 
red noise parametrized as a power law, $P(f) = A_\mathrm{red} f^{-\gamma}$, 
for those pulsars that show evidence of red noise 
(see Table~\ref{tab:sim} for more details). 
To these data sets we injected an isotropic stochastic GWB 
\sv{insert strength of GWB... are we going to have one value or more than one?}.
\begin{table}[htb]
	\setlength{\tabcolsep}{5pt}
	\caption{Pulsar parameters used in simulated PTAs.}
	\begin{center}
	\begin{tabular}{ccccc}
		\hline\hline
    		Pulsar	& $T_\mathrm{obs}$ (yrs) & $\sigma_w$ ($\mu\mathrm{s}$) & $A_\mathrm{red}$ & $\gamma_\mathrm{red}$ \\
		\hline
		J0030+0451 & 11.0 & & -13.93 & 3.56 \\
		J0613$-$0200 & 11.0 &  & -13.14 & 1.22 \\
		J1012+5307 & 11.0 &  & -12.79 & 1.51 \\
		J1024$-$0719 & 6.0 & & $-$ & $-$ \\
		J1455$-$3330 & 11.0 & & $-$ & $-$ \\
		J1600$-$3053 & 8.0 & & $-$ & $-$ \\
		J1614$-$2230 & 7.0 & & $-$ & $-$ \\
		J1640+2224 & 11.0 & & $-$ & $-$ \\
		J1713+0747 & 11.0 & & -14.14 & 1.58 \\
		J1741+1351 & 6.0 & & $-$ & $-$ \\
		J1744$-$1134 & 11.0 & & $-$ & $-$ \\
		B1855+09 & 11.0 & & -13.75 & 3.54 \\
		J1853+1303 & 7.0 & & $-$ & $-$ \\
		J1909$-$3744 & 11.0 &  & -13.84 & 1.74  \\
		J1918$-$0642 & 11.0 & & $-$ & $-$ \\
		J2010$-$1323 & 7.0 & & $-$ & $-$ \\
		J2145$-$0750 & 11.0 &  & -12.69 & 1.30 \\
		J2317+1439 & 11.0 & & $-$ & $-$ \\
    		\hline\hline
	\end{tabular}
	\end{center}
	\label{tab:sim}
\end{table}

@@

In Fig.~\ref{fig:os_datasetstats} we compare the fixed-noise values of $\hat{A}^2$ 
to the mean $\hat{A}^2$ for 200 different realizations of a GWB with $\Agw = 5\times10^{-15}$. 
The fixed-noise analysis tends to underestimate the strength of the stochastic background, 
whereas the noise-marginalized analysis is better able to recover the injected value. 
The mean value of the optimal statistic from the fixed-noise analysis, 
averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
while the mean value from the noise-marginalized analysis is $\hat{A}^2 = 2.5 \times 10^{-29}$. 
The noise-marginalized analysis also gives a higher average mean SNR of 1.7, 
compared to an average SNR of 1.0 using the fixed-noise analysis.
%\begin{figure}[ht]
%	\includegraphics[width=0.9\columnwidth]{plots/os_datasetstats_5e-15.pdf}
%	\caption{Optimal statistic and SNR for 425 simulated datasets 
%			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
%			We compare the values found using a fixed-noise analysis (blue) to the 
%			mean optimal statistic and mean SNR found by marginalizing over 
%			the individual pulsars' red noise parameters. The dashed vertical line 
%			indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
%			The value of $\hat{A}^2$ from the fixed-noise analysis, 
%			averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
%			while the noise-marginalized analysis gives a mean value of $\hat{A}^2 = 2.5 \times 10^{-29}$. 
%			Furthermore, the fixed-noise analysis gives an average SNR of 1.0, while the 
%			noise-marginalized analysis gives an average mean SNR of 1.7.}
%	\label{fig:os_datasetstats}
%\end{figure}
